Manuscript Figures

### 2021-03-11 src/manuscript_figures.Rmd
only print figures to a word document - provide urls to high-res images

---
title: "Pan-cancer analysis"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
theme_set(ggpubr::theme_pubclean())

# todo: move this to settings file
inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
assay_dir <- settings$assay_output$folder
surv <- do.call(rbind, readRDS(settings$surv))
projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
ggsave(paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf'), 
              plot = Fig, units = 'in', width = 12, height = 9)

@@

# clustering metrics of various cancers

Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs 
```{r clustering_metrics, fig.width=10, fig.height=8}
multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      M <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
    } 
    M <- t(subset_features_by_variance(t(M), topN = 10))

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig

ggsave(paste0('Figure', fig_count, '.cluster_stability_metrics.pdf'), 
       plot = Fig, units = 'in', 
       width = 10, height = 8)
```

# survival-predictive factors

I look for survival-predictive maui LFs, show how much c-index improves on top of 
clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.

```{r survival_factors, fig.width=12, fig.height=8}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

@@

fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf'), 
                         plot = Fig, width = 12, height = 14, units = 'in')

Figures[[fig_count]] <- Fig
```

# clinical factor associated MAUI LFs

Find maui LFs associated to clinical factors and plot them

```{r clinical_factors, fig.width=12, fig.height=9}
experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
# find clinical-factor associated LFs 
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('experiments', 'LFs', 'settings'))
clin.stats <- do.call(rbind, 
        pbapply::pbsapply(cl = cl, simplify = F, 
               experiments, function(experiment) {
               require(data.table)
               source(settings$utility_script)
               M <- LFs$maui_factors[[experiment]]
               clin <- process_clinical_covariates(rownames(M))
               dt <- do.call(rbind, 
                    pbapply::pbsapply(simplify = F, 
                                      colnames(clin), function(x) {
                      if(!is.numeric(clin[,x])) {
                        factors <- clin[,x]
                        m <- M[!is.na(factors),]
                        factors <- factors[!is.na(factors)]
                        dt <- get_factor_specific_variables(m, factors) 
                        dt$label <- x
                        return(dt)
                      }
                      return(NULL)
                      }))
               dt$project <- unlist(strsplit(experiment, "\\."))[[1]]
               return(dt)
             }))
parallel::stopCluster(cl)

@@

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))
Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))
fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.clinical_factors.LFs.pdf'), 
       plot = Fig, units = 'in', width = 12, height = 9)
Figures[[fig_count]] <- Fig
```

# Predicting MSI status using latent factors
```{r msi_stats, fig.width=10, fig.height=8}
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- settings$GDCdataDir
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))

@@

Fig <- cowplot::plot_grid(p1, 
                          plot_grid(p2.1, p2.2, p3, nrow = 1, labels = c('B', 'C', 'D')),
                          nrow = 2, labels = c('A', ''), rel_heights = c(5,3))
fig_count <- fig_count + 1
ggsave(filename = paste0("Figure", fig_count, '.MSI.status.pdf'), 
                         plot = Fig, units = 'in', width = 10, height = 8)

Figures[[fig_count]] <- Fig
```

# Semi-supervised clustering of samples

1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
2. Annotate each cluster using cancer state scores or hallmark gene set scores

## TCGA subtype informed clusters

### Use-case: pan-gastrointestinal cancers
```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
require(survminer)
require(cluster)
pr <- 'pan-gastrointestinal' 
experiment <- paste0(pr, '.cnv_gex_meth_mut')
M <- LFs$maui_factors[[experiment]]
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# find subtype specific LFs 
dt <- get_factor_specific_variables(M, st)

@@

fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf'), 
       Fig, units = 'in', 
       width = 10, height = 12)

Figures[[fig_count]] <- Fig
```

## Clustering informed by survival outcomes

Pick survival-predictive LFs and cluster samples based on those
```{r semi_supervised_survival, fig.width=10, fig.height=12}
# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

@@

fig_count <- fig_count + 1
ggsave(filename = paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf'), 
       plot = Fig, units = 'in', 
       width = 10, height = 12)

Figures[[fig_count]] <- Fig
```

>>>>

---
title: "Manuscript Figures"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE, warning = FALSE, message = FALSE)
# todo: pass this as argument
settings <- yaml::read_yaml('/data/local/buyar/arcas/pancancer_multiomics_manuscript/settings.yaml')
source(settings$utility_script)
library(data.table)
library(ggplot2)
library(pbapply)
library(ggrepel)
library(ggpubr)
library(cowplot)
theme_set(ggpubr::theme_pubclean())

# todo: move this to settings file
inputDir <- '/data/local/buyar/arcas/pancancer_multiomics_manuscript/output'
geneSetScoresFolder <- '/data/local/buyar/arcas/subtyping_paper/data/geneset_scores/singscore'
FigureFolder <- '/data/bimsbstatic/public/akalin/buyar/manuscript_figures_arcas'
assay_dir <- settings$assay_output$folder
surv <- do.call(rbind, readRDS(settings$surv))
projectDict <- sapply(yaml::read_yaml(settings$projectDict), 
                      function(x) gsub(' ', '', unlist(strsplit(x, ','))))
tcga_subtypes <- process_tcga_subtypes()

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.pancancer_cancer_type_prediction.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name), 
              plot = Fig, units = 'in', width = 12, height = 9)

@@

```{r clustering_metrics, fig.width=10, fig.height=8}
# clustering metrics of various cancers 
# Use top 10 factors (derived from multi-omics) to cluster samples using cancer-specific LFs

multi_omics_comb <- 'cnv_gex_meth_mut'
experiments <- grep(paste0('^TCGA-.*.', multi_omics_comb), names(LFs$pca_factors), value = T)
cl <- parallel::makeCluster(20)
parallel::clusterExport(cl, varlist = c('LFs', 'experiments', 'settings'))
clusters <- sapply(simplify = F, names(LFs), function(tool) {
  message("Clustering top ",tool)
  pbapply::pbsapply(cl = cl, simplify = F, experiments, function(experiment) {
    message(experiment)
    source(settings$utility_script)
    require(data.table)
    M <- LFs[[tool]][[experiment]]
    if(tool == 'maui_factors') {
      M <- remove_redundant_variables(M, perc = 99) # !!! make sure to use same setting as used for pan-cancer clustering
    } 
    M <- t(subset_features_by_variance(t(M), topN = 10))

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.cluster_stability_metrics.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(file.path(FigureFolder, fig.name),        
       plot = Fig, units = 'in', 
       width = 10, height = 8)
```

```{r survival_factors, fig.width=12, fig.height=8}
# survival-predictive factors
#I look for survival-predictive maui LFs, show how much c-index improves on top of 
#clinical factors (tumor stage/age/gender) and plot top survival-predictive factor per cancer.

# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.surv.Cindex_improv.hazard_ratio.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, width = 12, height = 14, units = 'in')
```

```{r clinical_factors, fig.width=12, fig.height=9}
# clinical factor associated MAUI LFs
# Find maui LFs associated to clinical factors and plot them

experiments <- grep('^TCGA-.*.cnv_gex_meth_mut', names(LFs$maui_factors), value = T)
# find clinical-factor associated LFs 
cl <- parallel::makeCluster(10)
parallel::clusterExport(cl, varlist = c('experiments', 'LFs', 'settings'))
clin.stats <- do.call(rbind, 
        pbapply::pbsapply(cl = cl, simplify = F, 
               experiments, function(experiment) {
               require(data.table)
               source(settings$utility_script)
               M <- LFs$maui_factors[[experiment]]
               clin <- process_clinical_covariates(rownames(M))
               dt <- do.call(rbind, 
                    pbapply::pbsapply(simplify = F, 
                                      colnames(clin), function(x) {
                      if(!is.numeric(clin[,x])) {
                        factors <- clin[,x]
                        m <- M[!is.na(factors),]
                        factors <- factors[!is.na(factors)]
                        dt <- get_factor_specific_variables(m, factors) 
                        dt$label <- x
                        return(dt)
                      }
                      return(NULL)
                      }))
               dt$project <- unlist(strsplit(experiment, "\\."))[[1]]
               return(dt)
             }))
parallel::stopCluster(cl)

@@

p2 <- cowplot::plot_grid(plotlist = plots[1:2], nrow = 2, labels = c('B', 'C'))

Fig <- cowplot::plot_grid(p1, p2, rel_widths = c(3,2), labels = c('A'))

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.clinical_factors.LFs.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 12, height = 9)
```

```{r msi_stats, fig.width=10, fig.height=8}
# Predicting MSI status using latent factors
# predict MSI status for pan-gastrointestinal tumors
# read MSI data
dataDir <- settings$GDCdataDir
MSI <- do.call(rbind, lapply(projectDict$`pan-gastrointestinal`, function(pr) {
  f <- file.path(dataDir, paste0(pr, '.MSI.tsv'))
  dt <- data.table::fread(f)[,c(2,4)]
  colnames(dt)[2] <- 'MSI'
  dt$project <- pr
  return(dt)
}))
# remove samples with undertermined MSI status
MSI <- MSI[MSI %in% c('MSI-H', 'MSI-L', 'MSS')]
msi_stats <- do.call(rbind, lapply(names(LFs), function(tool) {
  message(tool)
  M <- LFs[[tool]][['pan-gastrointestinal.cnv_gex_meth_mut']]
  y <- MSI[match(rownames(M), bcr_patient_barcode)]$MSI
  M <- M[!is.na(y),]
  y <- y[!is.na(y)]
  #res <- predict_cluster_labels(M, y, partition = 0.6, num.trees = 2000)
  res <- predict_cluster_labels_glmnet(M, y)
  #res <- predict_labels_glmnet(M, y)
  res$method <- tool
  return(res)
}))
msi_stats$method <- toupper(gsub('_factors', '', msi_stats$method))

@@

Fig <- cowplot::plot_grid(p1, 
                          plot_grid(p2.1, p2.2, p3, nrow = 1, labels = c('B', 'C', 'D')),
                          nrow = 2, labels = c('A', ''), rel_heights = c(5,3))
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.MSI.status.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', width = 10, height = 8)
```

```{r semi_supervised_subtypes, fig.width=10, fig.height=12}
# Semi-supervised clustering of samples 
# 1. Cluster samples by LFs associated to known TCGA subtypes or those that are associated to survival outcomes. 
#2. Annotate each cluster using cancer state scores or hallmark gene set scores

## TCGA subtype informed clusters 
## Use-case: pan-gastrointestinal cancers
require(survminer)
require(cluster)
pr <- 'pan-gastrointestinal' 
experiment <- paste0(pr, '.cnv_gex_meth_mut')
M <- LFs$maui_factors[[experiment]]
st <- tcga_subtypes[match(rownames(M), bcr_patient_barcode)]$Subtype_Selected
# find subtype specific LFs 
dt <- get_factor_specific_variables(M, st)

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.semi_supervised.clustering.', experiment,'.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       Fig, units = 'in', 
       width = 10, height = 12)
```

```{r semi_supervised_survival, fig.width=10, fig.height=12}

## Clustering informed by survival outcomes
## Pick survival-predictive LFs and cluster samples based on those

# factors to use in the model
clin_factors <- c('age_at_initial_pathologic_diagnosis', 'gender', 'ajcc_pathologic_tumor_stage')

@@

fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.semi_supervised.prognostic.clustering.', experiment,'.pdf')
names(Figures[fig.count]) <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', 
       width = 10, height = 12)

```

```{r print_figures, results='asis', include=TRUE, fig.width=10, fig.height=11}
base_url <- 'https://bimsbstatic.mdc-berlin.de'
for(i in 1:length(Figures)) {
  fig.url <- gsub("/data/bimsbstatic/public", base_url, file.path(FigureFolder, names(Figures)[i]))
  cat("# Figure",i)
  print(Figures[[i]])
  cat("High-res image:",fig.url,"\n\n")
}
```
